Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-05 analysis/paper/paper.qmd
work on Materials and Methods

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009). Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

@@

# Mateirals and Methods
To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007]. The remains considered in this study are from dwelling spaces and trash pits that reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded. These 40 units were selected because of their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices over time on a settlement scale. Moreover, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above provides a robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the arrival of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008]. For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions. For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using the Number of Identified Specimens (NISP) and classified into three groups: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., cranial parts and antlers) [@Madrigal2004]. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified. The proportion of deer bones with cut marks was calculated using the normed NISP for comparison between temporal sequences [@Grayson2004]. In addition, cut marks associated with the skinning process were examined. The skinning process can be identified by the encircling marks left around the lower limb, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

>>>>

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009). Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

@@

# Materials and Methods
To explore changes in dietary practices before and after the European arrival with associated long-distance trade, I examined faunal assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007]. The remains considered in this study are from dwelling spaces and trash pits that could reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded. These 40 units were selected because of their location that were believed to be the center of the settlement with thick archaeological deposits and relatively intact archaeological contexts. Also, the spatial continuity can provides a more consistent framework for understanding food practices over time on a settlement scale. The same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the average number of skeletal elements of that given taxa.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004]. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain. Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
